---
title: The chronoverse
subtitle: Towards a common language for chronological modelling in R
author: "Joe Roe<br><small><<https://joeroe.io>></small>"
affiliation: "University of Copenhagen"
url: "https://joeroe.io"
format: 
  revealjs:
    theme: ./gruvbox-dark.scss
    center: true
    echo: true
execute:
  cache: true
render:
  embed-resources: true
---

```{r echo=FALSE}
library(tibble)
options(pillar.print_min = 3)
options(pillar.print_max = 5)
options(pillar.width = 60)
```

```{r echo=FALSE, cache=FALSE}
library(conflicted)
conflicted::conflicts_prefer(c14::c14_calibrate)
conflicted::conflicts_prefer(c14::cal_sum)
conflicted::conflicts_prefer(dplyr::filter)
```

## 'Analysis of Dates and Chronological Patterns' in R {.small}

<small>Marwick et al., *CTV Archaeology*, <https://github.com/benmarwick/ctv-archaeology></small>

:::: {.columns}

::: {.column}
<small>Radiometric Dating</small>
: <small>rcarbon, rintcal, Bchron, clam, nimbleCarbon, rbacon, coffee, oxcAAR, ArchaeoPhases, spDates, IsoplotR, [c14]{.mark}, c14bazAAR, UThwigl, ADMUR, ArchaeoChron<br><br></small>

<small>Dendrochronology</small>
: <small>dendroNetwork, fellingdater</small>

<small>Luminescence Dating</small>
: <small>Luminescence, numOSL, BayLum</small>
:::

:::{.column}
<small>Paleoenvironmental Proxies</small>
: <small>tidypaleo, shoredate</small>

<small>Archaeological Time Series</small>
: <small>[era]{.mark}, aion, aoristic, kairos, aoristAAR, SPARTAAS, datplot, archSeries, eratosthenes</small>

<small>Stratigraphic Analysis</small>
: <small>archeofrag, [stratigraphr]{.mark}</small>

<small>Radiocarbon Datasets and APIs</small>
: <small>BSDA, c14bazAAR, p3k14c, xronos, rintchron, ArchaeoData</small>
:::

::::

## Idiomatic R

:::: {.fragments}

::: {.fragment .semi-fade-out}
* Vectorisation (everything is a vector)
* Rectangular data frames
* Functional programming
* Function-based OOP (S3, S7)
* Metaprogramming
:::

::: {.fragment}
* (Tidy) data as first argument
* Pipes for function composition (`%>%`, `|>`)
:::

::::

# 📦 c14

## 📦 c14 <<https://c14.joeroe.io>>

Tidy radiocarbon data:

* Converting and calibrating radiocarbon measurements
* S3 vector for calendar probability distributions <small>(`cal`, e.g. calibrated radiocarbon dates)</small>
* Summarising and aggregating calendar probability distributions
* Cleaning radiocarbon datasets <small>(e.g. `c14_control_lab_id()`)</small>
* Conversion between types in common radiocarbon packages

---

## Tidy radiocarbon calibration

```{r eg-c14-filter-mutate}
library(c14)
library(dplyr)
ppnd |>
  mutate(cal = c14_calibrate(cra, error)) |>
  filter(site == "Ganj Dareh") |>
  select(lab_id, cal)
```

---

## Tidy radiocarbon aggregation

```{r eg-c14-summarise}
ppnd |>
  mutate(cal = c14_calibrate(cra, error)) |>
  summarise(cal_sum = cal_sum(cal), .by = site) # or cal_density()
```

---

## `cal` vectors

vctrs-based S3 class for calendar probability distributions (e.g. calibrated radiocarbon dates)

```{r eg-cal}
ganj_dareh <- ppnd[ppnd$site == "Ganj Dareh",]
asiab <- ppnd[ppnd$site == "Asiab",]

ganj_dareh_cal <- c14_calibrate(ganj_dareh$cra, ganj_dareh$error)
asiab_cal <- c14_calibrate(asiab$cra, asiab$error)

c(ganj_dareh_cal, asiab_cal)
```

# 📦 stratigraphr

## <small>📦 stratigraphr <<https://stratigraphr.joeroe.io>></small>

Tidy framework archaeological stratigraphy and chronology:

* Read & write stratigraphic data
* Graph representation of stratigraphic sequences <small>(Dye & Buck 2015, <https://doi.org/10.1016/j.jas.2015.08.008>)</small>
* Validate and clean stratigraphs
* R interface to OxCal's Chronological Query Language (CQL)

---

## Import stratigraphic data

:::: {.columns}

::: {.column}
```{r eg-stratigraphr-read-csv}
library(readr)
read_csv("example_data/harris12.csv")
```
:::

::: {.column}
```{r eg-stratigraphr-read-lst}
library(stratigraphr)
read_lst("example_data/bonn.lst")
```
:::

::::

## Stratigraphs

```{r eg-stratigraphs}
h12 <- stratigraph(harris12, "context", "above")
strg_prune(h12) # Harris' 'Law of Succession'
```

## Plotting 'Harris matrixes'

E.g. with ggraph <<https://ggraph.data-imaginist.com/>>:

```{r eg-ggraph}
library(ggraph)
ggraph(h12, layout = "sugiyama") +
  geom_edge_elbow() +
  geom_node_label(aes(label = context), label.r = unit(0, "mm")) +
  theme_graph()
```

---

## Validating stratigraphs

```{r}
strat_is_valid(h12)
```

```{r warning=TRUE}
bad_h12 <- tidygraph::bind_edges(h12, data.frame(from = 6, to = 5))
strat_is_valid(bad_h12)
```

## Bayesian stratigraphic models with stratigraphr + c14

:::: {.columns}

::: {.column}
```{r}
shub1_radiocarbon |> 
  filter(!outlier) |> 
  group_by(phase) |> 
  summarise(
    cql = cql_phase(phase, 
                    cql_r_date(lab_id, cra, error))
  ) |> 
  arrange(desc(phase)) |> 
  summarise(
    cql = cql_sequence("Shubayqa 1", 
                       cql, 
                       boundaries = TRUE)
  ) |>
  pluck("cql") |>
  cql() ->
  shub1_cql
```
:::

::: {.column}
```{r echo=FALSE}
shub1_cql
```
:::

::::

---

📦 oxcAAR <<https://github.com/ISAAKiel/oxcAAR>>

```{r eg-oxcaar, eval=FALSE}
# Not executed (too slow)
library(oxcAAR)
quickSetupOxcal(path = fs::path_temp())
shub1_oxcal <- executeOxcalScript(shub1_cql)
readOxcalOutput(shub1_oxcal) |> 
  parseOxcalOutput() |> 
  plot()
```

# 📦 era

## 📦 era <<https://era.joeroe.io>>

* S3 vector class for year-based real time scales <small>(`yr`)</small>
* Standard for defining *eras* <small>(e.g. Common Era, Before Present, Hijri)</small>
* General function for conversion between eras
* Principled concatenation, coercion, and arithmetic rules
* Chronology-aware methods <small>(e.g. `yr_earliest()`, `yr_sort()`)</small>

## Years with era

```{r eg-yr}
library(era)
years <- yr(c(1100, 1200, 1300), "cal BP")
years
```

```{r eg-yr_transform}
yr_transform(years, "AH")
```

## Era definition

:::: {.columns}

::: {.column}
* label
* epoch
* name
* unit
* scale
* direction
:::

::: {.column}
```{r eg-era-def}
era(
    label = "T.A.", 
    epoch = -9021, 
    name = "Third Age", 
    unit = era_year("Gregorian"),
    scale = 1,
    direction = 1
)
```
:::

::::

## When it's working, you don't notice it

```{r eg-era-errors, error=TRUE}
x <- yr(1:3, "BP")
y <- yr(4:6, "cal BP")
z <- yr(7:9, "uncal BP")

y - x

c(y, z)

yr_transform(z, yr_era(x)) # No shortcut to calibration!
```

## 📦 tempo 

<small><https://github.com/joeroe/tempo></small>

* Temporal intervals (periods)
* Temporal relations (Allen + Levy's extended typology)

```{r eg-tempo}
library("tempo")
#> Loading required package: rlang

period1 <- c(1500, 1900)
period2 <- c(1800, 1950)

contemporary_with(period1, period2)
```

# A 'chronoverse'?

## An 'ecosystem' of packages

Packages that work well together *or* alone:

* Build up on base R (vector) types and generics
* Use explicit classes (S3, S4, or S7)
* Use understandable, predictable and unambiguous function names
* Have a common and predictable function signature (e.g. `.data`-first)
* Avoid complex/idiosyncratic data structures

## R package ecosystems

* tidyverse
* tidymodels
* easystats
* r-spatial / rspatial
* tesselle

## Rudiments of a 'chronoverse'

c14, stratigraphr, era:  
<small>(And tempo, rintchron, xronos, chronologr)</small>

* *vctrs*-based formalisms to represent different types of chronological vectors
* Functional paradigm for composing programs
* Assume/encourage the use of a tidy data

## <https://github.com/r-chrono> {.center}

# <small>The chronoverse – towards a common language for chronological modelling in R</small>

<small>Joe Roe <<https://joeroe.io>>  
University of Copenhagen</small>

:::: {.columns}

::: {.column}
📦 c14
: <small><https://c14.joeroe.io></small>

📦 stratigraphr
: <small><https://stratigraphr.joeroe.io></small>

📦 era
: <small><https://era.joeroe.io></small>

📦 tempo (v0.0.9000!)
: <small><https://github.com/joeroe/tempo></small>
:::

::: {.column}
{{< fa brands github >}} these slides:
: <small><https://github.com/joeroe/caa25_chronoverse></small>
:::

::::
